\thetitle

### 2021-03-03 writing/challenges.tex
challenges, start of rational: rewordings

At first glance implementing the method of section~\ref{sec:why-rational}
seems straightforward. The rational programmer is a simple algorithm and testing
whether the rational programmer hypothesis holds amounts to running the
algorithm on a collection of scenarios.

Three challenges remain, however: comparing the effect of blame on the rational
programmer across three different mechanisms; the collection of a large number
of representative debugging scenarios; and a proper sampling strategy for the
huge space of possibilities. A coincidental challenge is the disparity of the
implementations of gradually typed languages. To eliminate this variable, this
authors use Racket, which is thus far the only language in which all three major
soundness mechanisms are available in a reasonably robust and comparable manner;
for the implementation of the transient semantics, see the parallel submission
of an experience report~\cite{ttt21}.

The {\em first challenge\/} stems from the differences between the blame
assignment mechanisms of the three semantic variants.  While Natural assigns
blame to {\em one\/} component, Transient assigns blame to a sequence of
components. The Erasure semantics does not blame components {\it per se\/}. But,
it comes with an exception location and a stack trace, which implicitly suggests
parties to blame.  Each strategy affects the behavior of the rational programmer
differently (and real ones, too).

One way to reconcile these differences is to {\em equip the rational
programmer with modes \/} that represent the different types of
information the rational programmer takes into account when debugging a
scenario. Intuitively, with the common vocabulary of modes, different
blame strategies become different ways the rational programmer  interprets
blame.  For instance, one Transient mode may pick the oldest element of a
blame set as the next component to add types because it corresponds to the
earliest point in the program's evaluation that can discover an impedance
mismatch.  Another mode may opt to treat the sequence as a stack and add
types to its newest element.  When both modes are equally successful in
locating an impedance mismatch, measuring the rational programmer's effort
may answer which mode is the most effective.

However, attributing the success of the rational programmer to this or
that blame mechanism demands a careful look at the interplay between blame
and the run-time checks of each gradual type system. When a check fails
the Natural and Transient semantics assign blame and discard  the
information in the exception from the run-time check. Nevertheless, the
exception information may be equally or even more useful to the rational
programmer than blame. If this is the case blame may not play a critical
role for the rational programmer. Indeed, precisely because they do not
account for confounding factors, ~\citet{lksfd-popl-2020} cannot draw any
conclusions about the pragmatics of blame despite advertisements for the
opposite. Their experiment can only conclude that so-called blame-shifting
works but not that this is because of blame.

Modes offer a uniform way to compare the different semantics and  isolate
blame from the effect of the semantics's runtime checks. Specifically, the Natural and
Transient rational programmer come with a blame mode and an exception
mode. Only if the former succeeds debugging a program while the latter
fails, it is safe to conclude that blame is indeed necessary for the
rational programmer. Put differently, the exception mode serves as the
{\em baseline\/} for blame's value within a given semantics; if the
programmer in this mode ``beats'' the  blame one, a blame assignment
mechanism might be useless.

Besides the run-time checks, pure luck might also amplify the usefulness
of blame. Thus a proper evaluation of blame needs to factor out any random
effects too. A random rational programmer mode provides yet another necessary
baseline using the same ingredients as for exceptions.

@@

The {\em third challenge\/} is the explosive number of debugging scenarios that
result from the combination of mutation-based scenario generation and mode-based
analysis. All three factors---three different gradual typing systems, the large
number of mutants, and the number of debugging modes---contribute possibly
useful experimental data in a multiplicative manner. Hence, carried out naively,
the experiment would demand an unacceptably large amount of computational
resources.  A practical execution has no option but to {\em sample the space of
scenarios\/}, carefully ensuring that it still produces reproducible
conclusions.

>>>>

While at first glance the method of section~\ref{sec:why-rational} seems straightforward to implement and test,
three challenges remain. The first is comparing the effect of blame on the rational
programmer across three different mechanisms; second is the collection of a large number
of representative debugging scenarios; and third is finding a proper sampling strategy for the
huge space of possibilities. A coincidental challenge is the disparity of the
implementations of gradually typed languages. To eliminate this variable, the
authors use Racket, which is thus far the only language in which all three major
soundness mechanisms are available in a reasonably robust and comparable manner;
for the implementation of the transient semantics, see the parallel submission
of an experience report~\cite{ttt21}.

The {\em first challenge\/} stems from the differences between the blame
assignment mechanisms of the three semantic variants.  While Natural assigns
blame to {\em one\/} component, Transient assigns blame to a sequence of
components. The Erasure semantics does not blame components {\it per se\/}, but
it comes with an exception location and a stack trace, which implicitly suggests
parties to blame.  Each strategy affects the behavior of the rational programmer
differently (and real ones, too).

One way to reconcile these differences is to {\em equip the rational
programmer with modes \/} that represent the different types of
information the rational programmer takes into account when debugging a
scenario. Intuitively, with the common vocabulary of modes, different
blame strategies become different ways the rational programmer  interprets
blame.  For instance, one Transient mode may pick the oldest element of a
blame set as the next component to add types because it corresponds to the
earliest point in the program's evaluation that can discover an impedance
mismatch.  Another mode may opt to treat the sequence as a stack and add
types to its newest element.  If both modes are equally successful in
locating an impedance mismatch, measuring the rational programmer's effort with each mode
may answer which is the most effective.

However, attributing the success of the rational programmer to this or
that blame mechanism demands a careful look at the interplay between blame
and the run-time checks of each gradual type system. When a check fails
the Natural and Transient semantics assign blame instead of using the
information in the exception from the run-time check. Nevertheless, the
exception information may be equally or even more useful to the rational
programmer than blame. If this is the case, then blame may not play a critical
role for the rational programmer. Indeed, precisely because they do not
account for confounding factors, \citet{lksfd-popl-2020} cannot draw any
conclusions about blame specifically despite advertisements for the
opposite. Their experiment can conclude that so-called blame-shifting
works but not that this is because of blame.

Modes offer a uniform way to compare the different semantics and  isolate
blame from the effect of the semantics' runtime checks. Specifically, the Natural and
Transient rational programmer come with a blame mode and an exception
mode. If the former succeeds debugging a program while the latter
fails, then it is safe to conclude that blame is indeed necessary for the
rational programmer. Put differently, the exception mode serves as the
{\em baseline\/} for blame's value within a given semantics; if the
programmer in this mode performs as well or better than the  blame one, a blame assignment
mechanism might be useless.

Besides the run-time checks, the usefulness of blame might be due to pure luck.
Indeed, a proper evaluation of blame needs to consider the role of chance in blame assignment, too.
A random rational programmer mode provides yet another necessary baseline using the same ingredients as for exceptions.

@@

The {\em third challenge\/} is the explosive number of debugging scenarios that
result from the combination of mutation-based scenario generation and mode-based
analysis. All three factors---three different gradual typing systems, the large
number of mutants, and the number of debugging modes---contribute possibly
useful experimental data in a multiplicative manner. Hence, carried out naively,
the experiment would demand an infeasible amount of computational
resources.  A practical execution has no option but to {\em sample the space of
scenarios\/}, carefully ensuring that it still has reproducible
conclusions.

### 2021-03-03 writing/rational-natural.tex
challenges, start of rational: rewordings

The Natural semantics assigns blame to exactly one boundary.  A blame assignment
has the following specific meaning: the typed component may make incorrect type
assumptions about the untyped component in its interface, or the correct
interface exposes a bug in the untyped component. Our setup rules out the first
alternative (but see section~\ref{sec:conclusion}), and therefore the rational
programmer extends the trail to a scenario that swaps out the untyped component
for its typed counterpart.

Here is a rigorous definition.
\begin{quote}
\it A \emph{Natural blame trail} is a sequence of scenarios $\conf_0,...\conf_n$ of
a program $\system$ such that for all $0 \leq i \leq n - 1$, $\conf_i \subset
\conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i =\{\blame{\system, \conf_i}\}$ where
$\blame{\system, \conf}$ denotes the component that $\conf$ blames.
\end{quote}

When the buggy untyped component of a program is replaced by the typed counterpart,
the type checker fails if this component causes the impedance mismatch. Hence, a
trail that ends at an ill-typed scenarios successfully pinpoints the location of
bugs. 
\begin{quote}
\it A Natural blame trail $\conf_0,...\conf_n$ in a lattice $\lattice{\system}$ is
\emph{successful} iff (the program for) its last scenario $\conf_n$ does not type check.  A Natural
blame trail $\conf_0,..,\conf_n$ in a lattice $\lattice{\system}$ is \emph{failing}
iff (the program for) $\conf_n$ type checks and the trail cannot be extended further.
\end{quote}
That is, failing Natural blame trails are those that have not reached a scenario
that reveals the bug at compile time and ends in a scenario that does not blame
an untyped component during program execution. Thus the rational programmer gets
no hints from the gradual type system on how to continue the search for the bug.

While a successful Natural blame trail indicates that it 
pays off to heed blame assignments during debugging the trail's root, it does not answer whether
blame is a critical piece of the rational programmer's process.  For instance,
typing the top of stack trace of a failed run time type check, dubbed the
location of the exception, might be as useful as typing the blamed one.

@@

The definition of success for a Natural exception trail follows that for
a Natural blame trail.
Together the definitions for the two modes allow the comparison of the usefulness of blame 
with that for mere exceptions for debugging a scenario in the context of Natural semantics.
\begin{quote}
\it 
  Given a program $\system$ and a root $\conf_0$ in $\lattice{\system}$,
  Natural blame is \emph{more useful} than Natural exceptions for
  debugging $\conf_0$ iff 
  the Natural blame trail 
  that starts at $\conf_0$ is successful while the Natural exception trail that
  starts at $\conf_0$ is failing.
\end{quote}

>>>>

The Natural semantics assigns blame to exactly one boundary.  A blame assignment
has the following specific meaning: the typed component may make incorrect type
assumptions about the untyped component in its interface, or the correct
interface exposes a bug in the untyped component (or its dependencies). Our setup rules out the first
alternative (but see section~\ref{sec:conclusion}), and therefore the rational
programmer extends the trail to a scenario that swaps out the untyped component
for its typed counterpart.

Formally,
\begin{quote}
\it A \emph{Natural blame trail} is a sequence of scenarios $\conf_0,...\conf_n$ of
a program $\system$ such that for all $0 \leq i \leq n - 1$, $\conf_i \subset
\conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i =\{\blame{\system, \conf_i}\}$, where
$\blame{\system, \conf}$ denotes the component that $\conf$ blames.
\end{quote}

When the buggy untyped component of a program is replaced by the typed counterpart,
the type checker fails because this component causes the impedance mismatch. Hence, a
trail that ends at an ill-typed scenario successfully pinpoints the location of
the bug. 
\begin{quote}
\it A Natural blame trail $\conf_0,...\conf_n$ in a lattice $\lattice{\system}$ is
\emph{successful} iff (the program for) its last scenario $\conf_n$ does not type check.  A Natural
blame trail $\conf_0,..,\conf_n$ in a lattice $\lattice{\system}$ is \emph{failing}
iff (the program for) $\conf_n$ type checks and the trail cannot be extended further.
\end{quote}
That is, failing Natural blame trails are those that end in a scenario that does not reveal the bug statically, yet also does not blame
an untyped component. Thus the rational programmer gets
no hints from the gradual type system on how to continue the search for the bug.

While a successful Natural blame trail indicates that it 
pays off to heed blame assignments during debugging the trail's root, it does not answer whether
blame is a critical piece of the rational programmer's process.  For instance,
typing the top of a failed run time type check's stack trace, dubbed the
location of the exception, might be as useful as typing the blamed one.

@@

The definition of success for a Natural exception trail follows that for
a Natural blame trail.
Together, the definitions for the two modes allow the comparison of the usefulness of blame 
with that of mere exceptions for debugging a scenario in the context of Natural semantics.
\begin{quote}
\it 
  Given a program $\system$ and a root $\conf_0$ in $\lattice{\system}$,
  Natural blame is \emph{more useful} than Natural exceptions for
  debugging $\conf_0$ iff 
  the Natural blame trail 
  that starts at $\conf_0$ is successful while the Natural exception trail that
  starts at $\conf_0$ is failing.
\end{quote}

### 2021-03-03 writing/rational-stuff.tex
challenges, start of rational: rewordings

The actions of the rational programmer create an ascending chain---dubbed a
\emph{trail}---in $\lattice{\system}$.  The starting scenario $\conf_0$ of a
trail---also referred to as the \emph{root} (of the trail)---is the initial
interesting debugging scenario.  Starting from such a root, the rational
programmer steps along a trail. If the (program for this) scenario type-checks,
the program is run until it raises a run-time error.  The rational programmer
uses the information in the error to decide which component to equip with
types. This choice constructs scenario $\conf_{i+1}$ from $\conf_{i}$ and thus
creates a trail. The trail's construction ends when it reaches a scenarios that
contain the mutant module because the type checker rejects its typed version
outright.  At this point, the source of the impedance mismatch is identified.

>>>>

The actions of the rational programmer create an ascending chain---dubbed a
\emph{trail}---in $\lattice{\system}$.  The starting scenario $\conf_0$ of a
trail---also referred to as the \emph{root} (of the trail)---is the initial
debugging scenario.  Starting from such a root, the rational
programmer steps along a trail. If the (program for this) scenario type-checks,
the rational programmer runs the program until it raises a run-time error.  The rational programmer
uses the information in the error to decide which component to equip with
types. This choice constructs scenario $\conf_{i+1}$ from $\conf_{i}$ and thus
creates a trail. The trail's construction ends when it reaches a scenario that
contains the mutant module because the type checker rejects its typed version
outright.  At this point, the source of the impedance mismatch is identified.

### 2021-03-03 writing/rational.tex
challenges, start of rational: rewordings

Hence the research problem is how to make the actions of a rational programmer
comparable across all semantics. Sections~\ref{sub:natural}
through~\ref{sub:erasure} describe the various mode of the rational programmer and explain
to what extent they are comparable. Section\ref{sub:experiment}
enhances comparability with the programmer-effort dimension. Finally, the last
subsection states the experimental questions and gives an overview of the
experimental process.

>>>>

Hence the research problem is how to make the modes of the rational programmer
comparable even when they correspond to different semantics and offer different kinds of information. Sections~\ref{sub:natural}
through~\ref{sub:erasure} describe the various modes of the rational programmer and explain
the extent to which they are comparable. Section~\ref{sub:experiment}
enhances their comparability with the programmer-effort dimension. Finally, the last
subsection states the experimental questions and gives an overview of the
experimental process.

### 2021-03-03 writing/why-rational-methodology.tex
challenges, start of rational: rewordings

\end{enumerate}
Unfortunately, seeing an algorithm is different from using the algorithm to conduct
data-gathering experiments. The specific challenges are spelled out in the next
section, and the following three sections explains ways of overcoming them.

>>>>

That said, using the method to conduct data-gathering experiments entails
several challenges. The specific challenges are spelled out in the next section,
and the following three sections explain ways of overcoming them.
